---
title: "Technical Validation"
author: "Joe Corra"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(tidyverse)
library(plotly)
# library(gt)
# path for local development
#tvd <- readRDS("scripts/analysis/data_paper/technical_validation_chemistry_data")

# path for knitting
tvd <- readRDS("scripts/analysis/data_paper/technical_validation_chemistry_data")
```

## Flags

```{r flags, echo=FALSE}

flags_table <- tribble(
  ~Condition, ~ Value, ~Flag,
 "< MDL", "MDL",	"ND",
   "MDL and < RL",	"Measured concentration",	"L",
  "Holding time violation",	"Measured concentration",	"H",
  "Shipping issue",	"Measured concentration",	"S"
) 

mdl_table <- tvd$detlimits %>% 
  mutate(mdl = as.character(mdl), 
         year = ifelse (year == "2020_2021", "2020 - 2021", year), 
         analyte_group = if_else(analyte_group == "organic", "organic carbon", analyte_group), 
         name = str_to_upper(name), 
         name = if_else(name == "NO2_3", "combined NO2 and NO3", name),
         lab = if_else(lab == "CIN", "Cincinnati", "Ada")) %>%
  # group_by(analyte_group, lab, name, units) %>%
  # summarize(mdl = str_c(year, mdl, sep = ": ", collapse = "; "), .groups = "drop") %>%
  # ungroup() %>%
 select('analyte group' = analyte_group, analyte = name, lab, filtered, preservative, 'holding time' = holding_time, MDL = mdl, 'reporting limit' = ql) %>%
  arrange('analyte group')

# Break out by analyte group for easier in-text coding
anion_flags <- tvd$flags %>% filter('analyte group' == "anions") 
nutrient_flags <- tvd$flags %>% filter('analyte group' == "nutrients") 
oc_flags <- tvd$flags %>% filter('analyte group' == "organic carbon")  
metal_flags <- tvd$flags %>% filter('analyte group' == "metals") 
pigment_flags <- tvd$flags %>% filter('analyte group' == "algal indicators") 

# Break out by key analytes for easier in-text coding
sulfate_flags <- anion_flags %>% filter(analyte == "so4") 
ammonium_flags <- nutrient_flags %>% filter(analyte == "nh4")
nitrite_nitrate_flags <- nutrient_flags %>% filter(analyte == "no2.3")
tn_flags  <- nutrient_flags %>% filter(analyte == "tn")
tp_flags <- nutrient_flags %>% filter(analyte == "tp")
orthophosphate_flags <- nutrient_flags %>% filter(analyte == "op")
calcium_flags <- metal_flags %>% filter(analyte == "ca")
magnesium_flags <- metal_flags %>% filter(analyte == "mg")
tocdoc_flags <- oc_flags %>% filter(analyte %in% c("toc", "doc"))

# kable(flags_table, caption = "Table ???: Flag Descriptions")
# kable(mdl_table, caption = "Table ???: Minimum Detect Limit (MDL) for Key Analytes")
# kable(anion_flags)  
# kable(nutrient_flags)
# kable(oc_flags)  
# kable(metal_flags)
# kable(pigment_flags) 
# kable(tocdoc_flags)

```



### Main Document
Two flags associated with detection limits and an additional two flags associated with sample handling are included in 7_site_data.csv (**Table 2**). Procedures varied across labs for defining reporting limits (RL) and we assumed equivalency between the terms “quantification limit” and “reporting limit” provided by the labs. If neither value was reported by the lab, we assumed the RL was equal to the lowest non-zero standard in the standard curve. In some cases, one field duplicate was flagged for a detection limit issue, but the other was not; in such instances, the aggregated value was not flagged. In cases where duplicates had different flags, the flag associated with the higher-concentration sample was applied to the aggregated value. For example, if one duplicate was flagged ND and one received an L flag, the aggregated value received an L flag. 

Metal analytes had the greatest incidence of ND and L flags (`r metal_flags %>% summarize(nd = (sum(ND)/sum(total)) * 100) %>% pull(nd) %>% round(0)` and `r metal_flags %>% summarize(l = (sum(L)/sum(total)) * 100) %>% pull(l) %>% round(0)`%, respectively), followed by anions (`r anion_flags %>% summarize(nd = (sum(ND)/sum(total)) * 100) %>% pull(nd) %>% round(0)` and `r anion_flags %>% summarize(l = (sum(L)/sum(total)) * 100) %>% pull(l) %>% round(0)`%, respectively) and nutrients (`r nutrient_flags %>% summarize(nd = (sum(ND)/sum(total)) * 100) %>% pull(nd) %>% round(0)` and `r nutrient_flags %>% summarize(l = (sum(L)/sum(total)) * 100) %>% pull(l) %>% round(0)`%, respectively). Total and dissolved organic carbon concentrations were typically well above detection limit, with only `r oc_flags %>% summarize(nd = sum(ND)) %>% pull(nd)` ND flag and `r oc_flags %>% summarize(l = sum(L)) %>% pull(l)` L flag. Similarly, lab based chlorophyll-*a* had only `r pigment_flags %>% filter(analyte == "chla") %>% summarize(nd = sum(ND)) %>% pull(nd)` ND flags and `r pigment_flags %>% filter(analyte == "chla") %>% summarize(l = sum(L)) %>% pull(l)` L flags, but `r pigment_flags %>% filter(analyte == "microcystin") %>% summarize(nd = sum(ND)) %>% pull(nd)` and `r pigment_flags %>% filter(analyte == "microcystin") %>% summarize(l = sum(L)) %>% pull(l)` microcystin samples received ND or L flags, respectively. See **SI text** for a more detailed discussion. 

Shipping issues were rare and affected only `r pigment_flags %>% filter(analyte == "chla") %>% summarize(s = sum(S)) %>% pull(s)` chlorophyll-*a* samples, `r anion_flags %>% summarize(s = sum(S)) %>% pull(s)` anion samples, `r nutrient_flags %>% summarize(s = sum(S)) %>% pull(s)` nutrient samples, and `r pigment_flags %>% filter(analyte == "microcystin") %>% summarize(s = sum(S)) %>% pull(s)` microcystin samples. Holding time violations were more numerous and were present for `r anion_flags %>% summarize(h = (sum(H)/sum(total)) * 100) %>% pull(h) %>% round(0)`% of anion samples, `r nutrient_flags %>% summarize(h = (sum(H)/sum(total)) * 100) %>% pull(h) %>% round(0)`% of nutrient samples, `r pigment_flags %>% filter(analyte == "chla") %>% summarize(h = (sum(H)/sum(total)) * 100) %>% pull(h) %>% round(0)`% of chlorophyll-*a* samples, `r pigment_flags %>% filter(analyte == "microcystin") %>% summarize(h = (sum(H)/sum(total)) * 100) %>% pull(h) %>% round(0)`% of microcystin samples and `r oc_flags %>% summarize(h = (sum(H)/sum(total)) * 100) %>% pull(h) %>% round(0)`% of organic carbon samples. Many of these were due to disruptions in laboratory operations during the COVID pandemic. 

### SI
The incidence of detection limit flags varied considerably among the 39 water chemistry analytes included in this study. Here we focus on the analytes of greatest interest within each analyte group (i.e., anions, nutrients, metals, organic carbon, algal indicators) (Table S1). Among anion samples, `r (sum(anion_flags$L)/sum(anion_flags$total) * 100) %>% round(0)` percent were assigned L flags (measured concentration exceeded the minimum detection limit (MDL) but was less than the reporting limit, Table 2), but only `r (sum(sulfate_flags$L)/sum(sulfate_flags$total) * 100) %>% round(0)`% of sulfate (SO~4~^2-^) samples have L flags. Among nutrient samples, total nitrogen (TN) had the lowest prevalence of L flags `r (sum(tn_flags$L)/sum(tn_flags$total) * 100) %>% round(1)`% and ammonium (NH~4~^+^) the highest (`r (sum(ammonium_flags$L)/sum(ammonium_flags$total) * 100) %>% round(1)`%). Among metal samples, `r (sum(metal_flags$L)/sum(metal_flags$total) * 100) %>% round(0)`% were assigned L flags, but only `r (sum(magnesium_flags$L)/sum(magnesium_flags$total) * 100) %>% round(0)`% of magnesium (Mg) samples and `r (sum(calcium_flags$L)/sum(calcium_flags$total) * 100) %>% round(1)`% of calcium (Ca) samples, the metals of primary interest in this study, had L flags. In addition, `r pigment_flags %>% filter(analyte == "chla") %>% summarize(l = (sum(L)/ sum(total)) * 100) %>% pull(l) %>% round(0)`% of chlorophyll-*a* samples and `r pigment_flags %>% filter(analyte == "microcystin") %>% summarize(l = (sum(L)/sum(total)) * 100) %>% pull(l) %>% round(0)`% of microcystin samples had L flags. Organic carbon samples (both total organic carbon (TOC) and dissolved organic carbon (DOC)) had a relatively low incidence of L flags, with `r (sum(oc_flags$L)/sum(oc_flags$total) * 100) %>% round(1)` percent of samples having L flags.

Similarly, the incidence of ND flags (if measured concentration was below the MDL) differed among some key analytes when compared to their broader analyte groups. ND flags occurred in `r (sum(sulfate_flags$ND)/sum(sulfate_flags$total) * 100) %>% round(0)`% of SO~4~^2-^ samples compared to `r (sum(anion_flags$ND)/sum(anion_flags$total) * 100) %>% round(0)`% of all anion samples. Among nutrient samples, `r (sum(tn_flags$ND)/sum(tn_flags$total) * 100) %>% round(1)`% of TN samples, `r (sum(tp_flags$ND)/sum(tp_flags$total) * 100) %>% round(0)`% of TP samples, `r (sum(nitrite_nitrate_flags$ND)/sum(nitrite_nitrate_flags$total) * 100) %>% round(0)`% of combined nitrite-nitrate (NO~2_3~) samples, `r (sum(ammonium_flags$ND)/sum(ammonium_flags$total) * 100) %>% round(0)`% of NH~4~^+^ samples, and `r (sum(orthophosphate_flags$ND)/sum(orthophosphate_flags$total) * 100) %>% round(1)`% of soluble reactive phosphorus (OP) samples had ND flags, compared to `r (sum(nutrient_flags$ND)/sum(nutrient_flags$total) * 100) %>% round(0)`% of nutrient samples overall. No Ca or Mg samples had ND flags, compared to `r (sum(metal_flags$ND)/sum(metal_flags$total) * 100) %>% round(0)`% of all metal analytes. Only `r (sum(oc_flags$ND)/sum(oc_flags$total) * 100) %>% round(1)`% of organic carbon samples had ND flags. ND flags were rare among chlorophyll-a samples (`r pigment_flags %>% filter(analyte == "chla") %>% summarize(nd = (sum(ND) / sum(total)) * 100) %>% round(1)`%) but more prevalent among microcystin samples (`r pigment_flags %>% filter(analyte == "microcystin") %>% summarize(nd = (sum(ND) / sum(total)) * 100) %>% round(0)`%). 

## Field Blanks and Duplicates

```{r blanks, echo=FALSE}

  
# functions to get key values
blank_above_mdl <- function(df, analyte_select)
{summarize(df %>% list_rbind() %>% filter(analyte %in% analyte_select), x = (sum(blank >= mdl, na.rm = TRUE) / sum(n())) * 100) %>%
    pull(x) %>% round(2)}

blank_below_mdl <- function(df, analyte_select)
{summarize(df %>% list_rbind() %>% filter(analyte %in% analyte_select), x = (sum(blank < mdl, na.rm = TRUE) / sum(n())) * 100) %>%
    pull(x) %>% round(2)}

get_blank_mean <- function(data, analyte_select) {
  blanks_mean <- list_rbind(data) %>%
    filter(analyte %in% analyte_select) %>%
    filter(blank >= mdl, mean_unknown >= ql) %>%
    pull(blank_prop) %>%
    mean(na.rm = TRUE) 
  
  return(blanks_mean)
}

# Break out by analyte for easier in-text coding
blanks_so4 <- (get_blank_mean(tvd$blanks, "so4") *100) %>% round(2)
blanks_nh4 <- (get_blank_mean(tvd$blanks, "nh4") *100) %>% round(2)
blanks_no2.3 <- (get_blank_mean(tvd$blanks, "no2_3") *100) %>% round(2)
blanks_tn <-( get_blank_mean(tvd$blanks, "tn") *100) %>% round(2)
blanks_tp <- (get_blank_mean(tvd$blanks, "tp") *100) %>% round(2)
blanks_op <- (get_blank_mean(tvd$blanks, "op") *100) %>% round(2)
blanks_ca <- (get_blank_mean(tvd$blanks, "ca") *100) %>% round(2)
blanks_mg <- (get_blank_mean(tvd$blanks, "mg") *100) %>% round(2)
blanks_oc <- (get_blank_mean(tvd$blanks, c("toc", "doc")) *100) %>% round(2)

# percent of blanks < reporting limit
blanks_ql <- tvd$blanks %>% list_rbind() 
blanks_ql <- (100 * mean(blanks_ql$blank < blanks_ql$ql, na.rm = TRUE)) %>% round(2) 

blanks_mdl <- tvd$blanks %>% list_rbind() 
blanks_mdl <- (100 * mean(blanks_mdl$blank < blanks_mdl$mdl, na.rm = TRUE)) %>% round(2) 

blank_op <- tvd$blanks %>% list_rbind() %>% 
  filter(analyte == "op", blank >= mdl, mean_unknown >= ql) %>% 
  pull(blank) %>% mean(na.rm = TRUE) %>% round(2)

blank_tp <- tvd$blanks %>% list_rbind() %>% 
  filter(analyte == "tp", blank >= mdl, mean_unknown >= ql) %>% 
  pull(blank) %>% mean(na.rm = TRUE) %>% round(2)

unk_op <- tvd$blanks %>% list_rbind() %>% 
  filter(analyte == "op", blank >= mdl, mean_unknown >= ql) %>% 
  pull(mean_unknown) %>% mean(na.rm = TRUE) %>% round(2)

unk_tp <- tvd$blanks %>% list_rbind() %>% 
  filter(analyte == "tp", blank >= mdl, mean_unknown >= ql) %>% 
  pull(mean_unknown) %>% mean(na.rm = TRUE) %>% round(2)

```



### Main Document
Across all analytes (**Table S1**), measured concentrations in field blanks were below the MDL and RL in `r blanks_mdl %>% round(0)`% and `r blanks_ql %>% round(0)`% of samples, respectively. See **SI Text** for a more detailed discussion. Field duplicates were compared to assess variability related to sample handling and analysis. Relative percent difference (RPD) between field duplicates was negatively related to the concentration, such that very small differences between duplicates with low concentrations led to very high RPDs. To avoid overstating variability when absolute differences were small, we established analyte-specific concentration thresholds below which we report absolute difference and above which we report RPD. Information on thresholds, RPD, and absolute differences can be found in **SI Text**.


### SI
#### Field Blanks
While field blanks were generated using deionized or distilled water, the measured concentrations of key analytes typically met or exceeded the MDL. `r if(blank_above_mdl(tvd$blanks, "tn") == 100 & blank_above_mdl(tvd$blanks, "tp") == 100 & blank_above_mdl(tvd$blanks, "nh4") >= 97 & blank_above_mdl(tvd$blanks, "no2_3") >= 95 & blank_above_mdl(tvd$blanks, "op") == 100){"Greater than 95% of nutrient blanks (TN, TP, NH4+, NO2_3, OP) exceeded the MDL."}` `r if(blank_above_mdl(tvd$blanks, c("ca", "mg", "so4")) == 100){"All Ca, Mg, and SO42- blanks "}` and `r blank_above_mdl(tvd$blanks, c("toc", "doc")) %>% round(0)` percent of TOC and DOC blanks were above the MDL. Across all analytes (**Table S1**), measured concentrations in field blanks were below the MDL in `r blanks_mdl %>% round(0)`% of samples, while `r blanks_ql %>% round(0)`% were below the reporting limit.

Most field blanks had analyte concentrations that exceeded the MDL, but the measured concentration was generally small when compared to the unknowns. Here we focus on samples where the concentration in the unknown exceeded the reporting limit. Sulfate, Ca, and Mg blanks were, on average, equivalent to `r paste0(blanks_so4 %>% round(1), ", ", blanks_ca %>% round(1), ", and ", blanks_mg %>% round(1))`% of the concentrations in the unknown samples. Ammonium, NO~2_3~, and TN concentrations in field blanks were equivalent to `r paste0(blanks_nh4 %>% round(0), ", ", blanks_no2.3 %>% round(0), ", and ", blanks_tn %>% round(0))`% of those in the unknowns. Concentration of phosphorus in field blanks were equivalent to a larger fraction of those in the unknown as compared to other analytes (e.g., TP blanks equivalent to `r blanks_tp`% of unknown samples, OP blanks equivalent to `r blanks_op %>% round(0)`% of unknowns), but concentrations were quite low in both blank (mean TP = `r blank_tp %>% round(1)` ug L-1, mean OP = `r blank_op %>% round(1)` ug L-1) and unknown samples (mean TP = `r unk_tp %>% round(1)` ug L-1, mean OP = `r unk_op %>% round(1)` ug L-1). Measured concentrations of organic carbon blanks were `r blanks_oc %>% round(0)`% of unknown samples.

#### Field Duplicates

```{r dupes, echo=FALSE}



# Analyte,	Concentration Threshold, Mean RPD above threshold,	Mean absolute difference below threshold,	Units 
# Bind rows and create column with analyte group 

# Thresholds: # sulfate 24 mg/l    tn tp & no2_3 250 ug/l  
# ca mg 2.6 mg/l   op & ammonium 26 ug/l         

dupes <- tvd$dupes %>%
  group_by(lake_id, site_id, sample_depth, name, visit) %>%
  # calculate abs difference and rel. % difference for each set
  summarize(ad = abs(diff(value)), # difference between dup and unknown
            meanconc = mean(value, na.rm = TRUE), # mean of dup and unknown
            rpd = (ad/meanconc)*100) %>%
  ungroup() %>%
  # Add the thresholds for key analytes
    mutate(threshold = case_when(
    name == "so4" ~ 24, 
    name %in% c("ca", "mg") ~ 2.6, 
    name %in% c("tn", "tp", "no2_3")  ~ 250, 
    name %in% c("op", "nh4") ~ 26,
    .default = NA)
  ) %>%
  group_by(name) %>%
  # Get mean RPD and AD for these analytes, above or below thresholds
  # compute mean RPD for samples where mean conc > threshold
  summarize(mean_rpd_above_th = mean(rpd[meanconc >= threshold], na.rm = TRUE) %>% 
              round(2), 
              # compute mean AD for samples where mean conc < threshold
            mean_ad_below_th = mean(ad[meanconc < threshold], na.rm = TRUE) %>% 
              round(2), 
            # Mean RPD for all other analytes
            rpd = mean(rpd, na.rm = TRUE) %>% round(2), 
            threshold = min(threshold)) %>%
  ungroup() %>%
  # Add units   
  left_join(tvd$detlimits %>% 
              distinct(name, units), 
            by = "name")

# Two tables

# First table: analytes w/ thresholds
kable(dupes %>%
        filter(!is.na(threshold)) %>%
        select(analyte = name, 
               threshold, 
               'Mean RPD above threshold' = mean_rpd_above_th, 
               'Mean absolute difference below threshold' = mean_ad_below_th, 
               units))
# Second table: analytes w/out thresholds
kable(dupes %>%
        filter(is.na(threshold)) %>%
        select(analyte = name, 
               RPD = rpd, 
               units))
     
```

